Hazardous asteroids forecast via Markov random fields

### 2021-12-12 Presentation/Presentation.tex
PUB

\mode<presentation>
{
  \usetheme{Madrid}      % or try Darmstadt, Madrid, Warsaw, ...
  \usecolortheme{beaver} % or try albatross, beaver, crane, ...
  \usefonttheme{default}  % or try serif, structurebold, ...
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{caption}[numbered]
}

\usepackage[bookmarks]{hyperref}
\usepackage[backend=bibtex]{biblatex}
\usepackage{braket}
\addbibresource{bibliography.bib}
\newtheorem*{remark}{Remark}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{Pic/Intro.jpg}}
\title[Probabilistic modeling project]{Hazardous asteroids forecast via Markov random fields}
\subtitle{Project for the exam: Probabilistic Modelling (DSE)}
\author{Marzio De Corato}
\date{\today}

@@

\begin{frame}{Graphical models}
\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
\begin{equation}
X  \perp Y| Z \iff  p(X,Y|Z) = p(X|Z)p(Y|Z)
\end{equation}
\begin{equation}
p(\textbf{x}_{1:V})=p(x_{1})\prod^{V}_{t=1}p(x_{t}|x_{t-1})
\end{equation}
\end{frame}

\begin{frame}{Graphical models}
\begin{theorem}[Hammersley-Clifford]
A positive distribution $p(\textbf{y})>0$ satisfies the CI properties of an indirect graph G iif p can be represented as a product of factor, one per maximal clique,  i.e.
\begin{equation}
p(\textbf{y}|\theta)= \dfrac{1}{Z(\theta)}\prod_{c \in C }\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
where C is the set of all the (maximal) cliques of G,  and Z($\theta$) is the partition function given by 
\begin{equation}
Z(\theta):= \sum_{y}\prod_{c\in C}\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
Note that this partition function is what ensures the overall distribution sums to 1
\end{theorem}
\end{frame}

\begin{frame}{Graphical models}
\begin{equation}
p(y|\theta)=\dfrac{1}{Z(\theta)} exp\left(-\sum_{c}E(y_{c}|\theta_{c})\right)
\end{equation}
\begin{equation}
\psi_{c}(y_{c}|\theta_{c})=exp\left(-E(y_{c}|\theta_{c})\right)
\end{equation}
\end{frame}

\begin{frame}{Parameters tuning}
\begin{equation}
\frac{\partial\mathcal{L}}{\partial\boldsymbol{\theta}_{c}}=\frac{1}{N}\sum_{i}\left[\phi_{c}(y_{i})-\frac{\partial}{\partial\boldsymbol{\theta}_{c}}\log Z(\boldsymbol{\theta})\right]
\end{equation}

@@

\begin{frame}{Parameters tuning}
\begin{equation}
p\left(\textbf{y}|\boldsymbol{\theta}\right) = \dfrac{1}{Z(\theta)}\exp \left( \sum_{c}\boldsymbol{\theta}^{T}_{c}\phi_{c}\left(\textbf{y}\right)\right)
\end{equation}
\begin{equation}
\mathcal{L}\left(\boldsymbol{\theta}\right):= \frac{1}{N}\sum_{i}\log p\left(\textbf{y}_{i}|\boldsymbol{\theta}\right)=\frac{1}{N}\sum_{i}\left[\sum_{c} \boldsymbol{\theta}^{T}_{c}\phi_{c}(y_{i})-\log Z\left(\boldsymbol{\theta}\right)\right]
\end{equation}
\end{frame}

\begin{frame}{Gaussian graphical model}
\begin{equation}
f_{Y}(y)=\dfrac{det(K)^{1/2}}{(2\pi)^{d/2}}\exp\left(-\dfrac{1}{2}\textbf{y}^{T}K\textbf{y}\right)
\end{equation}

\begin{equation}
K=\Sigma^{-1}
\end{equation}

@@

\end{frame}
\begin{frame}{mmod}
\begin{center}
Decomposable model
\end{center}
\begin{equation}
\hat{f}(x)=\prod_{j=1}^{k}\dfrac{\hat{f}_{C_{j}}(x_{C}_{j})}{\hat{f}_{S_{j}}(x_{S}_{j})} 
\end{equation}
Maximized likehood
\begin{equation}
\hat{\mathcal{L}_{s}}=\sum_{i}n(i)log\left(\dfrac{n(i)}{N}\right)-Nq\dfrac{log(2\pi)}{2}-N\dfrac{log(det(S))}{2}-N\frac{q}{2}
\end{equation}

>>>>

\mode<presentation>
{
  \usetheme{Madrid}      % or try Darmstadt, Madrid, Warsaw, ...
  \usecolortheme{beaver} % or try albatross, beaver, crane, ...
  \usefonttheme{default}  % or try serif, structurebold, ...
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{caption}[numbered]
} 
\usepackage{pifont}
\usepackage[bookmarks]{hyperref}
\usepackage[backend=bibtex]{biblatex}
\usepackage{braket}
\addbibresource{bibliography.bib}
\newtheorem*{remark}{Remark}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{Pic/Intro.jpg}}
\title[Probabilistic modeling project]{Hazardous asteroids forecast via Markov random fields}
\subtitle{Project for the exam: Probabilistic Modelling (DSE)}
\author{Marzio De Corato}
\date{\today}

@@

\begin{frame}{Graphical models}
\begin{center}
\textbf{Problem:} How one can represent in a compact and elegant way the joint distribution $p(\textbf{x}|\boldsymbol{\theta})$ (where $\theta$ are the parameters)
\end{center}
\begin{center}
First one can consider to use the chain rule
\end{center}
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
\begin{center}
One can represent $p(x_{2},x_{1})$ as a $\mathcal{O}(K^{2})$ where the elements are $p(x_{2}=j|x_{1}=i)=T_{ij}$ and so on with high order tensors. T is called stochastic matrix
\end{center}
\end{frame}

\begin{frame}{Graphical models}

\begin{center}
In order to reduce the computational cost $\mathcal{O}(K^{2}V^{2})$ one can consider to take in to account the concept of conditional independence
\end{center}
\begin{equation}
X  \perp Y| Z \iff  p(X,Y|Z) = p(X|Z)p(Y|Z)
\end{equation}
\begin{center}
Therefore the distribution can be factorized as:
\end{center}
\begin{equation}
p(\textbf{x}_{1:V})=p(x_{1})\prod^{V}_{t=1}p(x_{t}|x_{t-1})
\end{equation}
\begin{center}
Which is a first order Markov chain
\end{center}
\end{frame}

\subsection{Markov random field}

\begin{frame}{}
\begin{center}
{\Huge Markov random field}
\end{center}
\end{frame}

\begin{frame}{Markov random field}
\begin{center}
\textbf{Problem:} differently with respect to Bayesian Network for an undirected graph there is no topological ordering associated to it. Therefore the chain rule cannot be used 
\end{center}
\begin{center}
\textbf{Solution:} One can use the potential function associated with maximal clique $\psi_{c}(\textbf{y}_{c}|\boldsymbol{\theta}_{c})$
\end{center}

\begin{frame}{Markov random field}
\begin{theorem}[Hammersley-Clifford]
A positive distribution $p(\textbf{y})>0$ satisfies the CI properties of an indirect graph G iif p can be represented as a product of factor, one per maximal clique,  i.e.
\begin{equation}
p(\textbf{y}|\boldsymbol{\theta})= \dfrac{1}{Z(\boldsymbol{\theta})}\prod_{c \in C }\psi_{c}(\textbf{y}_{c}|\boldsymbol{\theta}_{c})
\end{equation}
where C is the set of all the (maximal) cliques of G,  and Z($\boldsymbol{\theta}$) is the partition function given by 
\begin{equation}
Z(\boldsymbol{\theta}):= \sum_{y}\prod_{c\in C}\psi_{c}(\textbf{y}_{c}|\boldsymbol{\theta}_{c})
\end{equation}
Note that this partition function is what ensures the overall distribution sums to 1
\end{theorem}
\end{frame}

\begin{frame}{Markov random field}
\begin{center}
Gibbs distribution 
\end{center}
\begin{equation}
p(\textbf{y}|\boldsymbol{\theta})=\dfrac{1}{Z(\boldsymbol{\theta})}\exp\left(-\sum_{c}E(\textbf{y}|\boldsymbol{\theta})\right)
\end{equation}
\begin{center}
\begin{center}
Where $E(\textbf{y})$ is the energy associated with the variables in clique c. Therefore
\end{center}
\end{center}
\begin{equation}
\psi(\textbf{y}_{c}|\boldsymbol{\theta}_{c})=\exp\left(-E(\textbf{y}_{c}|\boldsymbol{\theta}_{c})\right)
\end{equation}

\begin{frame}{Markov random field}
\begin{equation}
\frac{\partial\mathcal{L}}{\partial\boldsymbol{\theta}_{c}}=\frac{1}{N}\sum_{i}\left[\phi_{c}(y_{i})-\frac{\partial}{\partial\boldsymbol{\theta}_{c}}\log Z(\boldsymbol{\theta})\right]
\end{equation}

@@

\begin{frame}{Markov random field}
\begin{equation}
p\left(\textbf{y}|\boldsymbol{\theta}\right) = \dfrac{1}{Z(\theta)}\exp \left( \sum_{c}\boldsymbol{\theta}^{T}_{c}\phi_{c}\left(\textbf{y}\right)\right)
\end{equation}
\begin{equation}
\mathcal{L}\left(\boldsymbol{\theta}\right):= \frac{1}{N}\sum_{i}\log p\left(\textbf{y}_{i}|\boldsymbol{\theta}\right)=\frac{1}{N}\sum_{i}\left[\sum_{c} \boldsymbol{\theta}^{T}_{c}\phi_{c}(y_{i})-\log Z\left(\boldsymbol{\theta}\right)\right]
\end{equation}
\end{frame}

\subsection{Graphical model algorithms}

\begin{frame}{}
\begin{center}
{\Huge Graphical model algorithms}
\end{center}
\end{frame}

\begin{frame}{Gaussian graphical model}
\begin{equation}
f_{Y}(y)=\dfrac{det(K)^{1/2}}{(2\pi)^{d/2}}\exp\left(-\dfrac{1}{2}\textbf{y}^{T}K\textbf{y}\right)
\end{equation}
\begin{equation}
K=\Sigma^{-1}
\end{equation}
\begin{equation}
\log L(K,\boldsymbol{\mu})=\frac{n}{2}\log\det(K)-\dfrac{n}{2}tr(KS)-\dfrac{n}{2}\left(\bar{\textbf{y}}-\boldsymbol{\mu}\right)^{T}\left(\bar{\textbf{y}}-\boldsymbol{\mu}\right)
\end{equation}

@@

\end{frame}
\begin{frame}{mmod}
\begin{center}
Decomposable model
\end{center}
\begin{equation}
\hat{f}(x)=\prod_{j=1}^{k}\dfrac{\hat{f}_{C_{j}}(x_{C}_{j})}{\hat{f}_{S_{j}}(x_{S}_{j})} 
\end{equation}
\begin{center}
Maximized likehood
\end{center}
\begin{equation}
\hat{\mathcal{L}_{s}}=\sum_{i}n(i)log\left(\dfrac{n(i)}{N}\right)-Nq\dfrac{log(2\pi)}{2}-N\dfrac{log(det(S))}{2}-N\frac{q}{2}
\end{equation}
